Accreditation: Post Shadbolt

### 2019-10-07 CEP20/CEP_accreditationpostshadbolt.tex
JHD edits

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
The promotion of quality via Accreditation by Professional, Statutory and Regulatory Bodies (PSRBs) is very much a feature of higher education in the united kingdom (UK).  In this changing  environment, there is a need for accreditation regimes to evolve in order to maximise the value they provide to Higher Education Institutions(HEIs).  This paper provides a summary in the context of one professional body (BCS, The Chartered Institute For IT) of 
\begin{enumerate}
\item what has happened in response to the recommendations of the Shadbolt review \cite{shadbolt2016shadbolt}
\item ongoing enhancement projects and the related future conversations that will help steer further enhancements
\end{enumerate}
\end{abstract}

@@

The Shadbolt Review \cite{shadbolt2016shadbolt} investigated high unemployment rates graduates of computer science and related degrees in the United Kingdom and the role of degree accreditation in promoting employability.The Shadbolt review noted that the rates of graduate unemployment were declining and varied considerably between geographic location and type of HEI. The discipline was already acting proactively to address the issue.  Related professional bodies have enhanced their accreditation processes in response these challenges.

BCS has made significant adjustments to their accreditation processes since 2015. These were partly in response to the recommendations of Shadbolt but also to other changes in the sector.  These have been communicated to the assessor community and to those responsible for leading the development of HEI accreditation applications. However, there are many academics who are less intimately involved in accreditation, this paper serves to communicate these and future enhancements to them and the wider computer science education community.

@@

Criticisms of the accreditation of degree programmes is not new, there is a history of claiming the processes are unnecessarily bureaucratic and constrain  innovation \cite{Harvey2004}. Equally its value has been highlighted particularly in the context of a potentially globally mobile workforce, and the dangers of accreditation streams being revenues streams in their own right rather than for the benefit of a discipline or wider society \cite{Knight_2015}.

@@

\subsubsection{Independent peer review}
Peer review is commonly used in quality regimes in higher education and in periodic review processes. In most cases the HEI being reviewed chooses the peer. As part of a BCS (or other professional body) a HEI has less choice regarding who completes a review and hence such a review is arguably more truly independent. BCS review panels contain a minimum of two experienced assessors who have strong awareness of the discipline norms across the UK sector.

@@

\subsubsection{Ensuring curricula relevance}
With the agreement of the sector (normally reached through the Council of Professors and Heads of Computing (CPHC)), it is possible to agree and promote curricula change. The most recent example of this is the inclusion of cybersecurity in all accredited degrees \cite{Cricketal2019}. BCS is in the process of mandating the inclusion of security in all accredited degree programmes to a standard agreed between professional bodies, industry and government. This is a work in progress \cite{Irons2016}. BCS has been requiring coverage since 2015 \cite[p.~17--18]{BCS2018a} with the result that all accredited universities must be compliant by 2020 (due to the five-year cycle). In the United States of America, the Association of Computing Machinery (ACM) has equally had cybersecurity (IAS: ``Information Assurance and Security'') in the curriculum since 2013~\cite{ACM2013a}, but it is not the accrediting body. The Accreditation Board for Engineering and Technology (ABET) is, and is requiring IAS with effect from the 2019-20 cycle (self-study reports due 1 July 2019): more precisely \cite[Table 3]{Oudshoornetal2018a} ``{\emph{The computing topics must include: \dots{} Principles and practices for secure computing\dots}}''. This will mean that  all accredited universities should be compliant by 2025 (due to their six-year cycle).

@@

In the United Kingdom the most common form of accreditation in the Computer Science and related discipline areas is by professional bodies.  The BCS, The Chartered Institute for IT (BCS) and the Institution of Engineering and Technology (IET) are the main bodies involved. The accreditation provided by these institutes are underpinned by international Initiatives such as the Washington Accord \cite{Washington2019} and Seoul Accord \cite{Seoul2019}. These memoranda support the internationalising of the curriculum and promote consistency and parity in Computer Science education globally. These professional bodies are also registered charities and hence have responsibilities for public good which extends beyond short term job market needs \cite{Stensaker2006}, \cite{Mutereko2017}. This is reflected in IETs and BCSs long history of expecting coverage environmental factors within the programmes they accredit and BCSs expectation of significant coverage of Legal, Ethical, Social and Professional Issues \cite{Brooke2018}.

@@

\section {Does it work?}	H
%how do you know? Give some evidence of effectiveness in context.

@@

\subsubsection{Not for Profit Organisations}
Tech Partnership Degrees is a Not For Profit Organisation that provides endorsements to Higher Education programmes with specific curricula elements aimed at job market requirements \cite{TP2019}. Tech Partnership Degrees have a specialist scope, endorsing programmes in the area of IT Management for Business and Software Engineering for Business.  Tech Partnership Degrees currently endorse 14 IT Management for Business Programmes and 5 Software Engineering for Business Programmes. As such Tech Partnership Degrees currently have limited impact upon more general Computer Science education. There is little published regarding the effectiveness or otherwise of this approach.

@@

\subsubsection{National Cyber Security Centre (NCSC)}
The NCSC is a UK Government organisation tasked with enhancing the cybersecurity of the UK. The NCSC publishes and accredits to a number of cybersecurity standards \cite{NCSC2018a}. These standard are linked to the Association of Computer Machinary (ACM) recommendations for curricula. To date the major focus of NSSC accreditation has been upon Masters degrees specializing in cybersecurity. More recently the NCSC has also began accrediting integrative masters programmes, undergraduate degrees in cybersecurity and Computer Science degrees with a significant cybersecurity focus \cite{NCSC2018b}.

@@

\subsubsection{The wider Computer Science education community}
The Computer Science discipline has responded actively to the challenges presented by Shadbolt with CPHC operating four working groups \cite{cphc_2016},  promotion of further engagement between industry and academia by initiatives such as the Royal Academy of Engineering Visiting Industrial Professor Scheme \cite{royal}

@@

\subsection{Internationalization}
BCS accreditation primarily takes place in the UK, however there are a small but growing number of institutions outside the UK that BCS accredits or who are seeking accreditation from BCS. Under the terms of the Washington and Seoul Accords, BCS does not engage in accreditation in the jurisdictions of Accord signatories without first consulting with the related local professional body.  On occasions BCS guidelines employ English / Welsh higher education terms without indicating that local equivalents are equally acceptable. The next set of guidelines will address this short coming.

\subsection {Engineering Council and Accreditation of Higher Education Programmes (AHEP) Version 4}
In the British jurisdiction of the Washington accord, the approach adopted is unusual in that the Engineering Council extends the license to accredit to a large number of professional bodies \cite{EC2019}. This is not the practice adopted in other jurisdictions, for example in the United States of America ABET is the sole body. A recent audit by the Washington Accord has highlighted divergence in practice in some areas with respect to Chartered Engineer Accreditation. The Engineering Council is in the process of seeking further consistency. This has lead to the new rules with respect of Compensation and Condonement for Eng accreditation \cite{EC2018}. Further changes are embedded in AHEP 4. The BCS requirements have been updated to reflect the expectations of Compensation and Condonement. Once AHEP 4 is finalised further work will be required to address these updates. Among other proposals, AHEP 4 intends to extend Chartered Engineering accreditation to include diversity and widening participation data (of staff and students) as part of the metrics that assess the quality of provision. The Computer Science discipline could clearly do better in this regard, so this should be a positive inclusion.

@@

Much has changed in BCS accreditation in the last few years. For many academics their only experience of BCS accreditation is the quinquennial BCS accreditation visit. Accreditation by the BCS is performed by panels of BCS assessors. An assessor may be an industrialist or an academic. The size of a panel varies depending upon the number of programmes a visited HEI is putting forward for consideration. A panel will always include an industrial assessor and two or more academic assessors.  At most HEIs will have two academic assessors.  There is a large pool of academic assessors and as such not all assessors will complete an accreditation visit every year. Not all HEIs have an academic assessor. Hence a significant set of the Computer Science education community are not informed regarding the evolution of accreditation practice.

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
The promotion of quality via Accreditation by Professional, Statutory and Regulatory Bodies (PSRBs) is very much a feature of higher education in the United Kingdom (UK).  In this changing  environment, there is a need for accreditation regimes to evolve in order to maximise the value they provide to Higher Education Institutions (HEIs).  This paper provides a summary in the context of one professional body (BCS, The Chartered Institute For IT) of 
\begin{enumerate}
\item what has happened in response to the recommendations of the Shadbolt review \cite{shadbolt2016shadbolt}
\item ongoing enhancement projects and the related future conversations that will help steer further enhancements
\end{enumerate}
\end{abstract}

@@

The Shadbolt Review \cite{shadbolt2016shadbolt} investigated the relatively high unemployment rates graduates of computer science and related degrees in the United Kingdom and the role of degree accreditation in promoting employability. The Shadbolt review noted that the rates of graduate unemployment were declining and varied considerably between geographic location and type of HEI. The discipline was already acting proactively to address the issue.  Related professional bodies have enhanced their accreditation processes in response these challenges.

BCS has made significant adjustments to their accreditation processes since 2015. These were partly in response to the recommendations of Shadbolt but also to other changes in the sector and the discipline.  These have been communicated to the assessor community and to those responsible for leading the development of HEI accreditation applications. However, there are many academics who are less intimately involved in accreditation, this paper serves to communicate these and future enhancements to them and the wider computer science education community.

@@

Criticisms of the accreditation of degree programmes is not new, there is a history of claiming the processes are unnecessarily bureaucratic and constrain  innovation \cite{Harvey2004},  and there are dangers of accreditation streams being revenues streams in their own right rather than for the benefit of a discipline or wider society \cite{Knight_2015}.
Equally its value has been highlighted particularly in the context of a potentially globally mobile workforce.

@@

\subsubsection{Independent peer review}
Peer review is commonly used in quality regimes in higher education and in periodic review processes. In most cases the HEI being reviewed chooses the peer. As part of a BCS (or other professional body) accreditation a HEI has less choice regarding who completes a review and hence such a review is arguably more truly independent. BCS review panels contain a minimum of two experienced assessors who have strong awareness of the discipline norms across the UK sector.

@@

\subsubsection{Ensuring curricula relevance}
With the agreement of the sector (normally reached through the Council of Professors and Heads of Computing (CPHC)), it is possible to agree and promote curricula change. The most recent example of this is the inclusion of cybersecurity in all accredited degrees \cite{Cricketal2019}. 
%BCS is in the process of mandating %% JHD strengthens text
BCS mandates the inclusion of security in all accredited degree programmes to a standard agreed between professional bodies, industry and government. This is a work in progress \cite{Irons2016}. BCS has been requiring coverage since 2015 \cite[p.~17--18]{BCS2018a} with the result that all accredited universities must be compliant by 2020 (due to the five-year cycle). In the United States of America, the Association of Computing Machinery (ACM) has equally had cybersecurity (IAS: ``Information Assurance and Security'') in the curriculum since 2013~\cite{ACM2013a}, but it is not the accrediting body. The Accreditation Board for Engineering and Technology (ABET) is, and is requiring IAS with effect from the 2019-20 cycle (self-study reports due 1 July 2019): more precisely \cite[Table 3]{Oudshoornetal2018a} ``{\emph{The computing topics must include: \dots{} Principles and practices for secure computing\dots}}''. This will mean that  all accredited universities should be compliant by 2025 (due to their six-year cycle).

@@

In the United Kingdom the most common form of accreditation in the Computer Science and related discipline areas is by professional bodies.  BCS, The Chartered Institute for IT (BCS) and the Institution of Engineering and Technology (IET) are the main bodies involved. The accreditation provided by these %institutes 
are underpinned by international Initiatives such as the Washington Accord \cite[for CEng]{Washington2019} and Seoul Accord \cite[for CITP]{Seoul2019}. These memoranda support the internationalising of the curriculum and promote consistency and parity in Computer Science education globally. These professional bodies are also registered charities and hence have responsibilities for public good which extends beyond short term job market needs \cite{Stensaker2006,Mutereko2017}. This is reflected in IETs and BCSs long history of expecting coverage environmental factors within the programmes they accredit and BCSs expectation of significant coverage of Legal, Ethical, Social and Professional Issues \cite{Brooke2018}.

@@

\section {Does it work?}
%how do you know? Give some evidence of effectiveness in context.

@@

\subsection{Not for Profit Organisations}
Tech Partnership Degrees is a Not For Profit Organisation that provides endorsements to Higher Education programmes with specific curricula elements aimed at job market requirements \cite{TP2019}. Tech Partnership Degrees have a specialist scope, endorsing programmes in the area of IT Management for Business and Software Engineering for Business.  Tech Partnership Degrees currently endorse 14 IT Management for Business Programmes and 5 Software Engineering for Business Programmes. As such Tech Partnership Degrees currently have limited impact upon more general Computer Science education. There is little published regarding the effectiveness or otherwise of this approach.

@@

\subsection{National Cyber Security Centre (NCSC)}
The NCSC is a UK Government organisation tasked with enhancing the cybersecurity of the UK. The NCSC publishes and accredits to a number of cybersecurity standards \cite{NCSC2018a}. These standard are linked to the Association of Computer Machinary (ACM) recommendations for curricula. To date the major focus of NSSC accreditation has been upon Masters degrees specializing in cybersecurity. More recently the NCSC has also began accrediting integrative masters programmes, undergraduate degrees in cybersecurity and Computer Science degrees with a significant cybersecurity focus \cite{NCSC2018b}.

@@

\subsection{The wider Computer Science education community}
The Computer Science discipline has responded actively to the challenges presented by Shadbolt with CPHC operating four working groups \cite{cphc_2016},  promotion of further engagement between industry and academia by initiatives such as the Royal Academy of Engineering Visiting Industrial Professor Scheme \cite{royal}

@@

\subsection{Internationalization}
BCS accreditation primarily takes place in the UK, however there are a small but growing number of institutions outside the UK that BCS accredits or who are seeking accreditation from BCS. Under the terms of the Washington and Seoul Accords, BCS does not engage in accreditation in the jurisdictions of Accord signatories without first consulting with the related local professional body.  On occasions BCS guidelines employ English / Welsh higher education terms without indicating that local equivalents are equally acceptable. The next set of guidelines will address this shortcoming.

\subsection {Engineering Council and Accreditation of Higher Education Programmes (AHEP) Version 4}
In the British jurisdiction of the Washington accord, the approach adopted is unusual in that the Engineering Council extends the license to accredit \emph{Chartered Engineer} to a large number of professional bodies \cite{EC2019}. This is not the practice adopted in other jurisdictions, for example in the United States of America ABET is the sole body. A recent audit by the Washington Accord has highlighted divergence in practice in some areas with respect to Chartered Engineer Accreditation. The Engineering Council is in the process of seeking further consistency. This has lead to the new rules with respect of Compensation and Condonement for Eng accreditation \cite{EC2018}. Further changes are embedded in AHEP 4. The BCS requirements have been updated to reflect the expectations of Compensation and Condonement. Once AHEP 4 is finalised further work will be required to address these updates. Among other proposals, AHEP 4 intends to extend Chartered Engineering accreditation to include diversity and widening participation data (of staff and students) as part of the metrics that assess the quality of provision. The Computer Science discipline could clearly do better in this regard, so this should be a positive inclusion.

@@

Much has changed in BCS accreditation in the last few years. For many academics their only experience of BCS accreditation is the quinquennial BCS accreditation visit. Accreditation by the BCS is performed by panels of BCS assessors. An assessor may be an industrialist or an academic. The size of a panel varies depending upon the number of programmes a visited HEI is putting forward for consideration. A panel will always include an industrial assessor and two or more academic assessors.  At most HEIs will have two academic assessors.  There is a large pool of academic assessors and as such not all assessors will complete an accreditation visit every year. Conversely, not all HEIs have an academic assessor. Hence a significant set of the Computer Science education community are not informed regarding the evolution of accreditation practice.
